
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
<!-- badges: end -->

<br>

# **crestr** A R package to perform probabilistic palaeoclimate reconstructions from fossil bioproxies

The goal of crestr is to produce probabilistic reconstructions of past  climate
change from fossil assemblage data. The approach is based on the estimation of
conditional responses of studies bioproxy studied to climate paramters. These
responses take the form of probability density functions (*pdfs*). The details
of the method have been described in [Chevalier *et al.* (2014)](http://www.doi.org/10.5194/cp-10-2081-2014)
and the calibrated data presented in [Chevalier (2019)](http://www.doi.org/10.1016/j.gloplacha.2019.01.016).

## Installation

You can install the development version from [GitHub](https://github.com/) with:


``` r
if(!require(devtools)) install.packages("devtools")
devtools::install_github("mchevalier2/crestr")
```

> **_NOTE:_**  This package is still in an **early phase of development**. The
documentation is sparse and the functionalities limited. If you notice any bug,
or if you would like to see some specific functions implemented, feel free to contact
me at <chevalier.manuel@gmail.com>. I will try to limit as much as possible
modifying existing functions and/or function parameters, but at the moment it
cannot be excluded. My apologies for this.

## Example

This is a basic example which shows you how to run crest with randomly generated
data:

```{r example}
library(crestr)
## loading example data
data(crest_ex)
data(crest_ex_pse)
data(crest_ex_selection)
```

Let's first have a look at the data. We have 20 fossil samples from which 7 taxa
have been identified. The data are already expressed in percentages.

```{r data_preview_fossil}
## the first 6 samples
head(crest_ex)
```

```{r data_preview_fossil2}
## the structure of the data frame
str(crest_ex)
```
