---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crestr

<!-- badges: start -->
<!-- badges: end -->

The goal of crestr is to produce probabilistic reconstructions of past climate
change from fossil assemblage data.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mchevalier2/crestr")
```
> **_NOTE:_**  This package is still in an **early phase of development**. The
documentation is sparse and the functionalities limited. If you notice any bug,
or if you would like to see some specific functions implemented, feel free to contact
me at <chevalier.manuel@gmail.com>. I will try to limit as much as possible
changing existing functions and/or function parameters, but at the moment this
cannot be excluded. My apologies for this.

## Example

This is a basic example which shows you how to run crest with randomly generated
data:

```{r example}
library(crestr)
## loading example data
data(crest_ex)
data(crest_ex_pse)
data(crest_ex_selection)
```

Let's first have a look at the data. We have 20 fossil samples from which 7 taxa
have been identified.

```{r data_preview_fossil}
## the first 6 samples
head(crest_ex)
```

```{r data_preview_fossil2}
## the structure of the data frame
str(crest_ex)
```


For each reconstruction, a proxy-species equivalency ('pse') table must be provided.
Here, with the 7 fake taxa, it looks like:
```{r data_preview_pse}
crest_ex_pse
```
Finally, one can specify which taxa should be used to reconstruct each variable:
```{r data_preview_selection}
crest_ex_selection
```

To illustrate the process, we will reconstruct bio1 (mean annual temperature)
and bio12 (annual precipitation) from these fake data. The description of the
different parameters is available in the first **vignette**.

```{r reconstructing_climate, results='hide'}
recons <- crest(
   df = crest_ex, pse = crest_ex_pse, taxaType = 0,
   climate = c("bio1", "bio12"), bin_width = c(2, 20),
   shape = c("normal", "lognormal"),
   selectedTaxa = crest_ex_selection, dbname = "crest_example"
)
```
Finally, the results can be visualised using the plot function and the
reconstructed climate values can be accessed from the nested `recons` object:

```{r extract_results_0}
names(recons)
lapply(recons$reconstructions, names)
```

```{r extract_results}
head(recons$reconstructions$bio1$optima)
str(recons$reconstructions$bio1$optima)
```

```{r extract_results2}
signif(recons$reconstructions$bio1$posterior[1:6, 1:6], 3)
str(recons$reconstructions$bio1$posterior)
```

```{r plot}
plot(recons, climate = 'bio1', plot = TRUE , save = FALSE)
plot(recons, climate = 'bio12')
```

## References

 - Chevalier, M., Cheddadi, R., Chase, B.M., 2014. CREST (Climate REconstruction SofTware): a probability density function (PDF)-based quantitative climate reconstruction method. *Clim. Past* 10, 2081–2098.
 - Chevalier, M., 2019. Enabling possibilities to quantify past climate from fossil assemblages at a global scale. *Glob. Planet. Change* 175, 27–35.
