---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crestr

<!-- badges: start -->
<!-- badges: end -->

The goal of crestr is to produce probabilistic reconstructions of past climate
change from fossil assemblage data.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mchevalier2/crestr")
```
## Example

This is a basic example which shows you how to run crest with randomly generated
data:

```{r example}
library(crestr)
## loading example data
data(crest_ex)
data(crest_ex_pse)
data(crest_ex_selection)
```

Let's first have a look at the data. We have 20 fossil samples from which 7 taxa
have been identified.

```{r data_preview_fossil}
## the first 6 samples
head(crest_ex)
```

```{r data_preview_fossil2}
## the structure of the data frame
str(crest_ex)
```


For ech reconstruction, a proxy-species equivalency ('pse') table must be provided.
Here, with the 7 fake taxa, it looks like:
```{r data_preview_pse}
crest_ex_pse
```
Finally, one can specify which taxa should be used to reconstruct each variable:
```{r data_preview_selection}
crest_ex_selection
```

To illustrate the process, we will reconstruct bio1 (mean annual temperature)
and bio12 (annual precipitation) from these fake data. The description of the
different parameters is available in the first *vignette*.

```{r reconstructing_climate, results='hide'}
recons <- crest(
   df = crest_ex, pse = crest_ex_pse, taxaType = 0,
   climate = c("bio1", "bio12"), bin_width = c(2, 20),
   shape = c("normal", "lognormal"),
   selectedTaxa = crest_ex_selection, dbname = "crest_example"
)
```
Finally, we can visualise the results using the plot function:

```{r plot}
plot(recons, climate = 'bio1', plot = TRUE , save = FALSE)
plot(recons, climate = 'bio12')
```

The reconstructed climate values can be accessed from the *recons* object:

```{r extract_results}
head(recons$reconstructions$bio1$optima)
```

```{r extract_results2}
recons$reconstructions$bio1$posterior[1:6, 1:10]
```
