---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crestr

<!-- badges: start -->
[![MIT license](https://img.shields.io/badge/License-MIT-blue.svg)](https://lbesson.mit-license.org/)
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active) <br >
[![codecov](https://codecov.io/gh/mchevalier2/crestr/branch/master/graph/badge.svg)](https://codecov.io/gh/mchevalier2/crestr)
[![Travis build status](https://travis-ci.com/mchevalier2/crestr.svg?branch=master)](https://travis-ci.com/mchevalier2/crestr)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/mchevalier2/crestr?branch=master&svg=true)](https://ci.appveyor.com/project/mchevalier2/crestr) <br >
[![CRAN status](https://www.r-pkg.org/badges/version/crestr)](https://CRAN.R-project.org/package=crestr)
![CRAN/METACRAN](https://img.shields.io/cran/v/crestr) <br >
[![GitHub R package version](https://img.shields.io/github/r-package/v/mchevalier2/crestr)](https://img.shields.io/github/r-package/v/mchevalier2/crestr)
[![HitCount](http://hits.dwyl.com/mchevalier2/crestr.svg)](http://hits.dwyl.com/mchevalier2/crestr)
[![Downloads](https://img.shields.io/github/downloads/mchevalier2/crestr/total)]()
[![Last commit](https://img.shields.io/github/last-commit/mchevalier2/crestr)]()
<!-- badges: end -->

<hr />

The goal of crestr is to produce probabilistic reconstructions of past climate
change from fossil assemblage data. The approach is based on the estimation of
conditional responses of studies bioproxy studied to climate paramters. These
responses take the form of probability density functions (*pdfs*). The details
of the method have been described in [Chevalier *et al.* (2014)](http://www.doi.org/10.5194/cp-10-2081-2014)
and the calibrated data presented in [Chevalier (2019)](http://www.doi.org/10.1016/j.gloplacha.2019.01.016).

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mchevalier2/crestr")
```

> **_NOTE:_**  If the install fails, it is possible that
[**gdal**](https://gdal.org/download.html) or [**PROJ**](https://proj.org/install.html)
are not installed on your system. Ample documentation exists online to guide you
through the installation of these two elements to your own system.


## contact

If you experience any trouble while using this package, or if you can think of
additional functionalities to incorporate to the package, please contact me at
<chevalier.manuel@gmail.com> or open a discussion [here](https://github.com/mchevalier2/crestr/issues).


## Example

The following example illustrates the basics of **crestr** using pseudo-data
(*i.e.* randomly generated data). More elaborate examples can be found at
<https://mchevalier2.github.io/crestr/>.


```{r example}
library(crestr)
## loading example data
data(crest_ex)
data(crest_ex_pse)
data(crest_ex_selection)
```

Let's first have a look at the data. The dataset is composed of 20 fossil
samples from which 7 taxa have been identified. The data are expressed in percentages.

```{r data_preview_fossil}
## the first 6 samples
head(crest_ex)
##
## the structure of the data frame
str(crest_ex)
```

For each reconstruction, a proxy-species equivalency ('pse') table must be provided.
Here, with the 7 pseudo-taxa, it looks like:
```{r data_preview_pse}
crest_ex_pse
```

Finally, unique sets of taxa can be specified to reconstruct each climate
variable. In the example, *bio1* (mean annual temperature) and *bio12* (annual
precipitation) will be reconstructed. The dataset has been designed so that
Taxa 1, 2, 3 and 7 are sensitive to *bio1* while Taxa 1, 4, 5 and 7 are sensitive
to *bio12*.

```{r data_preview_selection}
crest_ex_selection
```

These pseudo data can be provided to the **crest** function and provided some
parameters (see the full vignettes for a detail of these parameters), the
reconstructions will be processed.

```{r reconstructing_climate, results='hide'}
recons <- crest(
   df = crest_ex, pse = crest_ex_pse, taxaType = 0,
   climate = c("bio1", "bio12"), bin_width = c(2, 20),
   shape = c("normal", "lognormal"),
   selectedTaxa = crest_ex_selection, dbname = "crest_example"
)
```

The climate sampled by the data can be graphically represented for a quick
assessment of the calibration dataset.

```{r plot-climate-space}
plot_climateSpace(recons)
```

Additional graphical tools can be used to assess which taxa should/could be used
for each variable. On the following example, it is clear that Taxon2 has a much
stronger correlation with *bio1* than to *bio12*, hence its selection for *bio1*
only.

```{r plot-plot_taxaCharacteristics}
plot_taxaCharacteristics(recons, taxanames='Taxon2')

```

The results can be quickly visualised using the plot function and the
reconstructed climate values can be accessed from the nested `recons` object:

```{r extract_results_0}
names(recons)
lapply(recons$reconstructions, names)
```

```{r extract_results}
head(recons$reconstructions$bio1$optima)
str(recons$reconstructions$bio1$optima)
```

```{r extract_results2}
signif(recons$reconstructions$bio1$posterior[1:6, 1:6], 3)
str(recons$reconstructions$bio1$posterior)
```


```{r plot,  fig.show="hold", out.width="50%"}
plot(recons, climate = 'bio1')
plot(recons, climate = 'bio12', simplify=TRUE, uncertainties=c(0.4, 0.6, 0.8))
```

If satisfying, the results can be directly exported from the R environment in
unique spreadsheets for each variables (or csv files) and the **crest** object
is exported as an RData file to enable easy reuse in the future.

```{r export}
export(recons, loc=getwd(), dataname='crest-test')
list.files(file.path(getwd(), 'crest-test'))
```

```{r delete export example, echo = FALSE, results = "hide"}
unlink(file.path(getwd(), 'crest-test'), recursive=TRUE)
```

## References

 - Chevalier, M., Cheddadi, R., Chase, B.M., 2014. CREST (Climate REconstruction
   SofTware): a probability density function (PDF)-based quantitative climate
   reconstruction method. *Clim. Past* 10, 2081–2098. [10.5194/cp-10-2081-2014](http://www.doi.org/10.5194/cp-10-2081-2014)
 - Chevalier, M., 2019. Enabling possibilities to quantify past climate from
   fossil assemblages at a global scale. *Glob. Planet. Change* 175, 27–35. [10.1016/j.gloplacha.2019.01.016](http://www.doi.org/10.1016/j.gloplacha.2019.01.016)
