---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crestr

<!-- badges: start -->
<!-- badges: end -->

The goal of crestr is to produce probabilistic reconstructions of past climate
change from fossil assemblage data. The approach is based on the estimation of
conditional responses of studies bioproxy studied to climate paramters. These
responses take the form of probability density functions (*pdfs*). The details
of the method have been described in [Chevalier *et al.* (2014)](http://www.doi.org/10.5194/cp-10-2081-2014)
and the calibrated data presented in [Chevalier (2019)](http://www.doi.org/10.1016/j.gloplacha.2019.01.016).

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mchevalier2/crestr")
```
> **_NOTE:_**  This package is still in an **early phase of development**. The
documentation is sparse and the functionalities limited. If you notice any bug,
or if you would like to see some specific functions implemented, feel free to contact
me at <chevalier.manuel@gmail.com>. I will try to limit as much as possible
modifying existing functions and/or function parameters, but at the moment it
cannot be excluded. My apologies for this.

## Example

This is a basic example which shows you how to run crest with randomly generated
data:

```{r example}
library(crestr)
## loading example data
data(crest_ex)
data(crest_ex_pse)
data(crest_ex_selection)
```

Let's first have a look at the data. We have 20 fossil samples from which 7 taxa
have been identified. The data are already expressed in percentages.

```{r data_preview_fossil}
## the first 6 samples
head(crest_ex)
```

```{r data_preview_fossil2}
## the structure of the data frame
str(crest_ex)
```


For each reconstruction, a proxy-species equivalency ('pse') table must be provided.
Here, with the 7 fake taxa, it looks like:
```{r data_preview_pse}
crest_ex_pse
```
Finally, one can specify which taxa should be used to reconstruct each variable:
```{r data_preview_selection}
crest_ex_selection
```

To illustrate the process, we will reconstruct bio1 (mean annual temperature)
and bio12 (annual precipitation) from these fake data. The description of the 
different parameters is available in the first **vignette**.

```{r reconstructing_climate, results='hide'}
recons <- crest(
   df = crest_ex, pse = crest_ex_pse, taxaType = 0,
   climate = c("bio1", "bio12"), bin_width = c(2, 20),
   shape = c("normal", "lognormal"),
   selectedTaxa = crest_ex_selection, dbname = "crest_example"
)
```

The climate sampled by the data can be assessed graphically using the figure
```{r plot climate space}
plot_climateSpace(recons)

```


Finally, the results can be visualised using the plot function and the
reconstructed climate values can be accessed from the nested `recons` object:

```{r extract_results_0}
names(recons)
lapply(recons$reconstructions, names)
```

```{r extract_results}
head(recons$reconstructions$bio1$optima)
str(recons$reconstructions$bio1$optima)
```

```{r extract_results2}
signif(recons$reconstructions$bio1$posterior[1:6, 1:6], 3)
str(recons$reconstructions$bio1$posterior)
```

```{r plot}
plot(recons, climate = 'bio1', plot = TRUE , save = FALSE)
plot(recons, climate = 'bio12')
```

## References

 - Chevalier, M., Cheddadi, R., Chase, B.M., 2014. CREST (Climate REconstruction SofTware): a probability density function (PDF)-based quantitative climate reconstruction method. *Clim. Past* 10, 2081–2098. [10.5194/cp-10-2081-2014](http://www.doi.org/10.5194/cp-10-2081-2014)
 - Chevalier, M., 2019. Enabling possibilities to quantify past climate from fossil assemblages at a global scale. *Glob. Planet. Change* 175, 27–35. [10.1016/j.gloplacha.2019.01.016](http://www.doi.org/10.1016/j.gloplacha.2019.01.016)
