---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crestr

<!-- badges: start -->
[![MIT license](https://img.shields.io/badge/License-MIT-blue.svg)](https://lbesson.mit-license.org/)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://lifecycle.r-lib.org/articles/stages.html) <br >
[![CRAN status](https://www.r-pkg.org/badges/version/crestr)](https://CRAN.R-project.org/package=crestr)
![CRAN/METACRAN](https://img.shields.io/cran/v/crestr)
[![GitHub R package version](https://img.shields.io/github/r-package/v/mchevalier2/crestr)](https://img.shields.io/github/r-package/v/mchevalier2/crestr) <br >
<!-- badges: end -->

#

The goal of `crestr` is to enable probabilistic reconstructions of past climate
change from fossil assemblage data. The approach is based on the estimation of
conditional responses of different bio-proxies to various climate parameters.
These responses take the form of probability density functions (*PDFs*). The
details of the method have been described in
[Chevalier *et al.* (2014)](http://www.doi.org/10.5194/cp-10-2081-2014), the
calibrated data are presented in
[Chevalier (2019)](http://www.doi.org/10.1016/j.gloplacha.2019.01.016) and the package is fully described in [Chevalier (2022)](https://cp.copernicus.org/articles/18/821/2022/).


## Installation

You can install the package from [CRAN](https://CRAN.R-project.org/package=crestr).
``` r
install.packages("crestr")
```

You can also install the development version from [GitHub](https://github.com/) with:

``` r
if(!require(devtools)) install.packages("devtools")
devtools::install_github("mchevalier2/crestr")
```

> **_NOTE:_**  If the install fails, it is possible that
[**gdal**](https://gdal.org/download.html) or [**PROJ**](https://proj.org/install.html)
are not installed on your system. Ample documentation exists online to guide you
through the installation of these two elements to your own system.


## Contact

If you experience any trouble while using this package, or if you think of
additional functionalities to incorporate to the package, please contact me at
<chevalier.manuel@gmail.com> or open a discussion [here](https://github.com/mchevalier2/crestr/issues).


## How to use `crestr`

### Online documentation

The package is fully documented and the help of each function can be accessed
with `?function`. More detailed information as well as documented examples are
available from **<https://mchevalier2.github.io/crestr/>**.

### A Quick Example

The following example illustrates the basics of `crestr` using pseudo-data
(_i.e._ randomly generated data).


```{r example}
library(crestr)
## loading example data
data(crest_ex)
data(crest_ex_pse)
data(crest_ex_selection)
```

Let's first have a look at the data. The dataset is composed of 20 fossil
samples from which 7 taxa have been identified. The data are expressed in percentages.

```{r data_preview_fossil}
## the first 6 samples
head(crest_ex)
##
## the structure of the data frame
str(crest_ex)
```

For each reconstruction, a proxy-species equivalency ('pse') table must be provided.
Here, with the 7 pseudo-taxa, it looks like:
```{r data_preview_pse}
crest_ex_pse
```

Finally, unique sets of taxa can be specified to reconstruct each climate
variable. In the example, *bio1* (mean annual temperature) and *bio12* (annual
precipitation) will be reconstructed. The dataset has been designed so that
Taxa 1, 2, 3 and 7 are sensitive to *bio1* while Taxa 1, 4, 5 and 7 are sensitive
to *bio12*. Check <https://mchevalier2.github.io/crestr/articles/get-started.html>
for more details on this selection.

```{r data_preview_selection}
crest_ex_selection
```

These pseudo-data can be provided to the **crest** function and provided some
parameters (see the full vignettes for a detail of these parameters), the
reconstructions will be processed.

```{r reconstructing_climate, results='hide'}
recons <- crest(
   df = crest_ex, pse = crest_ex_pse, taxaType = 0,
   climate = c("bio1", "bio12"), bin_width = c(2, 50),
   shape = c("normal", "lognormal"),
   selectedTaxa = crest_ex_selection, dbname = "crest_example"
)
```

A specific print function was created to summarise the crestObj.

```{r showcase-print}
recons
```

The climate sampled by the data can be graphically represented for a quick
assessment of the calibration dataset.

```{r plot-climate-space, out.width="100%", fig.height=6, fig.width=6}
plot_climateSpace(recons)
```


Additional graphical tools can be used to assess which taxa should/could be used
for each variable. On the following example, it is clear that Taxon2 has a much
stronger correlation with *bio1* than to *bio12*, hence its selection for *bio1*
only.

```{r plot-plot_taxaCharacteristics, fig.width=7.5, fig.height=8.8}
plot_taxaCharacteristics(recons, taxanames='Taxon2')
```

The results can be quickly visualised using the plot function and the
reconstructed climate values can be accessed from the nested `recons` object:

```{r extract_results_0}
names(recons)
lapply(recons$reconstructions, names)
```

```{r extract_results}
head(recons$reconstructions$bio1$optima)
str(recons$reconstructions$bio1$optima)
```

```{r extract_results2}
signif(recons$reconstructions$bio1$likelihood[1:6, 1:6], 3)
str(recons$reconstructions$bio1$likelihood)
```


```{r plot,  fig.show="hold", out.width="50%"}
plot(recons, climate = 'bio1')
plot(recons, climate = 'bio12', simplify=TRUE, uncertainties=c(0.4, 0.6, 0.8))
```

If satisfying, the results can be directly exported from the R environment in
unique spreadsheets for each variables (or csv files) and the **crest** object
is exported as an RData file to enable easy reuse in the future.

```{r export}
export(recons, loc=getwd(), dataname='crest-test')
list.files(file.path(getwd(), 'crest-test'))
```

```{r delete export example, echo = FALSE, results = "hide"}
unlink(file.path(getwd(), 'crest-test'), recursive=TRUE)
```


## References

 - Chevalier, M., Cheddadi, R., Chase, B.M., 2014. CREST (Climate REconstruction
   SofTware): a probability density function (PDF)-based quantitative climate
   reconstruction method. *Clim. Past* 10, 2081–2098. [10.5194/cp-10-2081-2014](http://www.doi.org/10.5194/cp-10-2081-2014)
 - Chevalier, M., 2019. Enabling possibilities to quantify past climate from
   fossil assemblages at a global scale. *Glob. Planet. Change* 175, 27–35.
   [10.1016/j.gloplacha.2019.01.016](http://www.doi.org/10.1016/j.gloplacha.2019.01.016)
 - Chevalier, M., 2022. _crestr_ an R package to perform probabilistic climate
   reconstructions from palaeoecological datasets. Clim. Past
   [doi:10.5194/cp-18-821-2022](https://www.doi.org/10.5194/cp-18-821-2022)
